---
title: "Identification with a categorical treatment"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Identification with a categorical treatment}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(catIV)
```

## The problem

With a binary treatment, an instrument identifies the effect among
*compliers*: those who take the treatment when encouraged and not otherwise.
With three or more alternatives that description falls apart. A decision team
might be willing to prescribe A or B but never C. Another might consider B or
C but never A. An instrument that shifts the ranking of all three moves these
two groups in different directions, and "complier" no longer names a single
population.

`catIV` handles this by making the set of alternatives a decision team would
actually consider — its **adherence set** — the central object.

## A price-ranking instrument

Three treatments, and a procurement ranking that changes each year. Each
instrument value is the alternatives sorted from most to least encouraged,
which in this setting means cheapest first.

```{r design}
treatments <- c("cheap", "mid", "dear")

instrument <- list(
  "2019" = c("cheap", "mid",   "dear"),
  "2020" = c("mid",   "cheap", "dear"),
  "2021" = c("mid",   "dear",  "cheap"),
  "2022" = c("dear",  "mid",   "cheap")
)
```

The behavioural assumption, **categorical monotonicity**, is that a decision
team takes whichever alternative in its adherence set the instrument
encourages most:

```{r choice}
choose_treatment(instrument[["2019"]], c("mid", "dear"))
choose_treatment(instrument[["2022"]], c("mid", "dear"))
```

A team open to `mid` or `dear` takes `mid` in 2019, when `mid` is cheaper, and
`dear` in 2022, when the ranking flips. A team open only to `dear` takes
`dear` in both years and contributes nothing to identification.

## Enumerating adherence sets

With three alternatives there are `2^3 - 1 = 7` non-empty adherence sets.

```{r sets}
sets <- adherence_sets(treatments)
sets
```

The **response matrix** records what each set receives under each instrument
value. It is the whole behavioural model in one table.

```{r response}
R <- response_matrix(sets, instrument)
R
```

Rows `A1` to `A3` are constant down every column: a team considering only one
alternative always takes it, whatever the price. Those teams carry no
information. The variation in the other columns is what identifies effects.

## What is identifiable, and for whom?

From the response matrix, `projection_matrices()` builds the indicator matrix
`B_t` for each treatment and the projector `K_t` onto its null space. The
vectors `b` that identify an effect are the binary solutions of `b K_t = 0`.

```{r identify}
KB <- projection_matrices(R, treatments)
b  <- solve_b_pairs(KB)

target_populations(b)
```

All three pairwise contrasts are identified, but each for a **different**
population:

- `cheap` vs `mid` applies to teams whose adherence set contains both, that is
  `A4 = {cheap, mid}` or `A7 = {cheap, mid, dear}`.
- `cheap` vs `dear` applies only to teams whose set is *exactly*
  `A5 = {cheap, dear}`. Teams that would also consider `mid` are excluded,
  because for them a price change can divert choice to `mid` rather than
  between the two alternatives being compared.
- `mid` vs `dear` applies to teams containing both.

Reading this off is the point of the framework. An effect estimate means
little without knowing whose effect it is.

## Estimating

`simulate_adherence()` generates data by exactly this mechanism, so we know
the answer in advance. Treatment response probabilities are 0.20, 0.35 and
0.50, and the confounder shifts the outcome as well as the adherence set — so
a crude comparison will be biased.

```{r simulate}
set.seed(1)
d <- simulate_adherence(
  n = 40000, instrument = instrument, sets = sets, treatments = treatments,
  v_levels = c(0, 1), v_probs = c(0.5, 0.5),
  v_to_set = list("0" = c(3, 1, 1, 2, 4, 1, 3),
                  "1" = c(1, 3, 2, 1, 2, 3, 4)),
  v_on_y = c(0.10, 0.30), t_on_y = c(0.20, 0.35, 0.50)
)
head(d)
```

Estimation is two observable quantities and then algebra. `P_Z` is the
probability of each treatment at each instrument value:

```{r pz}
p_z <- estimate_p_z(d, "Z", "T", instrument_levels = rownames(R),
                    treatments = treatments)
round(p_z, 3)
```

Treatment tracks the price ranking, which is the instrument doing its work.
`Q_Z` combines outcome and treatment, and `estimate_p_sigma()` gives the size
of each target population:

```{r estimate}
q_z     <- estimate_q_z(d, "Z", "T", "Y", instrument_levels = rownames(R),
                        treatments = treatments)
p_sigma <- estimate_p_sigma(p_z, KB, b)

lapply(p_sigma$average, round, 3)
```

```{r late}
late <- estimate_late(q_z, KB, b, p_sigma)
lapply(late, round, 3)
```

True contrasts are `-0.15`, `-0.30` and `-0.15`. Compare with a crude
comparison of outcomes by treatment received:

```{r crude}
crude <- tapply(d$Y, d$T, mean)
round(c(cheap_mid  = crude[["cheap"]] - crude[["mid"]],
        cheap_dear = crude[["cheap"]] - crude[["dear"]],
        mid_dear   = crude[["mid"]]   - crude[["dear"]]), 3)
```

The crude estimates are pulled toward zero because teams inclined to the
dearer options also have better outcomes for reasons unrelated to treatment.

## Why not just compare two at a time?

A natural shortcut is to throw away everyone who took a third alternative,
collapse the instrument to "is A cheaper than B?", and run the usual Wald
ratio. `naive_iv()` does exactly that:

```{r naive}
round(c(
  cheap_mid  = naive_iv("cheap_mid",  d, instrument, "Z", "T", "Y"),
  cheap_dear = naive_iv("cheap_dear", d, instrument, "Z", "T", "Y"),
  mid_dear   = naive_iv("mid_dear",   d, instrument, "Z", "T", "Y")
), 3)
```

These are much better than the crude comparison, because the instrument still
removes the confounding. What they do not fix is *which* movement is being
measured. Making `cheap` cheaper does not only move teams between `cheap` and
`mid`; it also pulls in teams who would otherwise have taken `dear`. The
two-at-a-time analysis attributes that movement to the `cheap` versus `mid`
contrast, so the estimate is a blend of contrasts rather than the one asked
for.

How much that matters depends on the design. Here the effects are homogeneous
across decision teams, so blending contrasts is mild and the naive estimates
land within a few hundredths — the `mid` versus `dear` estimate is the most
visibly off. With effects that differ across alternatives or across teams the
discrepancy grows, and in either case the naive estimate has no clean
interpretation: there is no population for which it is the average effect.

## Restricting the adherence sets

Some adherence sets may be clinically implausible, and ruling them out is an
extra assumption. It changes what is identified and for whom — not always
helpfully. Here is the effect of dropping each set in turn:

```{r restricted}
describe <- function(keep) {
  tp <- target_populations(
    solve_b_pairs(projection_matrices(response_matrix(sets[keep], instrument),
                                      treatments))
  )
  vapply(tp, function(x) if (length(x) == 0) "none" else
    paste(names(x), collapse = " or "), character(1))
}

t(vapply(names(sets), function(drop) describe(setdiff(names(sets), drop)),
         character(3)))
```

Dropping `A5` removes the `cheap` vs `dear` contrast altogether: `A5` was the
only population for which it was identified. Dropping `A4` narrows `cheap` vs
`mid` from `A4 or A7` to `A7` alone. Restriction here costs information rather
than adding it, because with three treatments and four instrument values all
three contrasts are identified already.

The benefit appears with more alternatives, and it is **descriptive**. With
five treatments there are 31 adherence sets, and the target population of an
identified contrast can be a union of many of them — arithmetically fine,
impossible to describe to a clinician:

```{r five, eval = FALSE}
# The design in paper 2: five TNF inhibitors, ten procurement periods.
target_populations(solve_b_pairs(projection_matrices(R_full, TNFi)))$Cer_Inf
#> $`A12+A18+A23+A24+A27+A28+A30+A31`
#> [1] "A12" "A18" "A23" "A24" "A27" "A28" "A30" "A31"
```

Restricting to the 14 adherence sets clinicians considered realistic leaves
the same seven contrasts identified, but the target population for that
contrast becomes a single set — decision teams open to any of the five drugs.
A describable population is what makes the estimate usable, which is why the
restriction was worth the assumption.

## Next steps

- `vignette("target-trials")` covers covariate adjustment, bootstrap
  confidence intervals, and describing the target population as a trial.
- `vignette("heterogeneous-effects")` covers effects that vary with a
  covariate.
