---
title: "Adjustment, inference and target trials"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Adjustment, inference and target trials}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(catIV)
```

This vignette picks up where `vignette("identification")` leaves off: you have
an identified effect, and now you need a confidence interval for it and a way
to say who it applies to. Both matter more here than in a randomised trial,
because the target population is not the enrolled sample but an unobservable
subgroup defined by adherence sets.

## Setup

```{r design}
treatments <- c("cheap", "mid", "dear")
instrument <- list(
  "2019" = c("cheap", "mid",   "dear"),
  "2020" = c("mid",   "cheap", "dear"),
  "2021" = c("mid",   "dear",  "cheap"),
  "2022" = c("dear",  "mid",   "cheap")
)

sets <- adherence_sets(treatments)
R    <- response_matrix(sets, instrument)
KB   <- projection_matrices(R, treatments)
b    <- solve_b_pairs(KB)
```

Simulated data with a measured baseline covariate `age` that predicts both
treatment and outcome:

```{r data}
set.seed(2)
d <- simulate_adherence(
  n = 6000, instrument = instrument, sets = sets, treatments = treatments,
  v_levels = c(0, 1), v_probs = c(0.5, 0.5),
  v_to_set = list("0" = c(3, 1, 1, 2, 4, 1, 3),
                  "1" = c(1, 3, 2, 1, 2, 3, 4)),
  v_on_y = c(0.10, 0.30), t_on_y = c(0.20, 0.35, 0.50)
)
d$age <- round(50 + 8 * (d$A %in% c("A3", "A6", "A7")) + rnorm(nrow(d), 0, 6))
```

## Covariate adjustment

The unadjusted estimators use raw cell proportions. The adjusted versions fit
a model and then **standardise** over the observed covariate distribution at
each instrument value, which is what keeps them valid inputs to the
identification algebra. A per-covariate-profile prediction would not be.

```{r adjusted}
p_z_adj <- estimate_p_z(d, "Z", "T", covariates = "age", parametric = TRUE,
                        instrument_levels = rownames(R),
                        treatments = treatments)
q_z_adj <- estimate_q_z(d, "Z", "T", "Y", covariates = "age",
                        parametric = TRUE, family = "binomial",
                        p_z = p_z_adj, instrument_levels = rownames(R),
                        treatments = treatments)

p_sigma_adj <- estimate_p_sigma(p_z_adj, KB, b)
lapply(estimate_late(q_z_adj, KB, b, p_sigma_adj), round, 3)
```

Note that adjustment is refused unless you ask for it explicitly:

```{r refuse, error = TRUE}
estimate_p_z(d, "Z", "T", covariates = "age")
```

Because the instrument is as good as randomly assigned here, adjustment buys
precision rather than correcting bias. It matters when the instrument is only
conditionally valid.

## Confidence intervals

There is no closed-form variance for these ratio estimators, so inference is
by bootstrap. `bootstrap_late()` resamples, re-estimates everything on each
resample, and takes empirical quantiles.

```{r boot}
bs <- bootstrap_late(300, d, "Z", "T", "Y", KB, b, n_cores = 2)
round(bs$ci, 3)
```

Three things worth knowing:

- **`n_cores` is a speed setting, nothing more.** Each replicate draws from
  its own random number stream, so the same `seed` gives the same intervals
  however many workers you use, and on whatever machine. It defaults to one
  fewer than the cores available to your session.
- **`n_kept` is not decoration.** A resample that loses an instrument value
  cannot be estimated and is recorded as `NA`. If `n_kept` is well below the
  number of replicates, the interval rests on less than it appears to.
- **`cap = TRUE`** discards replicates whose estimate falls outside the range
  of the outcome. When a target population is small, the denominator of the
  ratio occasionally lands near zero and throws out an absurd value that would
  otherwise drag the quantiles.

```{r multiple-imputation, eval = FALSE}
# With multiply imputed data, pass a list. Each imputation contributes n
# replicates and all of them are pooled before quantiles are taken.
bootstrap_late(300, list(imp1, imp2, imp3), "Z", "T", "Y", KB, b)
```

## Describing the target population

The awkward part of reporting these estimates is saying who they apply to.
"Decision teams whose adherence set contains both alternatives" is correct but
not a sentence a clinician can act on. The target trial framing helps: if the
adherence set *were* observable, you could run a trial that enrolled the teams
in the target population and randomised them among the alternatives they were
willing to accept. The IV analysis estimates the result of that trial.

`pseudo_population()` builds the weighted dataset corresponding to that trial.
Give it the **unadjusted** `p_sigma`: the weights are built from the raw data,
so mixing in a standardised `P_Sigma` breaks the correspondence.

```{r pseudo}
p_z     <- estimate_p_z(d, "Z", "T", instrument_levels = rownames(R),
                        treatments = treatments)
q_z     <- estimate_q_z(d, "Z", "T", "Y", instrument_levels = rownames(R),
                        treatments = treatments)
p_sigma <- estimate_p_sigma(p_z, KB, b)

pseudo <- pseudo_population("cheap_mid", d, KB, b, p_sigma, "Z", "T")
nrow(pseudo)
range(pseudo$w)
```

Weights can be negative. `b B_t^+` is a contrast across instrument values, not
a probability, so some observations enter with a negative sign. That is
expected rather than a numerical failure.

Weighting makes an ordinary comparison of outcomes reproduce the IV estimate:

```{r pseudo-check}
arm1 <- pseudo$T == "cheap"
weighted.mean(pseudo$Y[arm1], pseudo$w[arm1]) -
  weighted.mean(pseudo$Y[!arm1], pseudo$w[!arm1])

estimate_late(q_z, KB, b, p_sigma)$cheap_mid
```

The point of the pseudo-population is not the effect estimate, though — it is
the baseline table. Weighted covariate means describe the population the
effect applies to, in the same way a trial's Table 1 describes who was
enrolled:

```{r baseline}
tapply(seq_len(nrow(pseudo)), pseudo$T, function(i) {
  round(weighted.mean(pseudo$age[i], pseudo$w[i]), 1)
})

# Unweighted, for comparison
round(tapply(d$age, d$T, mean)[c("cheap", "mid")], 1)
```

The weights average to exactly one within each treatment arm, so
`mean(x * w)` within an arm equals the weighted mean of `x` and either form
gives the same table.

```{r weights-average-one}
tapply(pseudo$w, pseudo$T, mean)
```

## Choosing what to report

With several identified contrasts, not all deserve reporting. A contrast whose
target population is tiny will have a wide interval and describe almost
nobody. `estimate_p_sigma()` gives the sizes, and screening on them is how
paper 2 chose which comparisons to present:

```{r screen}
sizes <- vapply(p_sigma_adj$average, max, numeric(1))
round(sizes, 3)

names(sizes)[sizes > 0.1]
```
